\ \\ \LARGE\bf Speculative Evaluation in Particle Swarm Optimization \thanks{Matthew Gardner, Andrew McNabb, and Kevin Seppi are with the Department of Computer Science, Brigham Young University, 3361 TMCB, Provo, UT 84602 (phone: 801-422-8717; email: \{mjg82,a,k\}@cs.byu.edu).}

### 2010-04-21 honors_thesis/journal_paper/paper.tex
Updating the results section.  Getting close to done

@@

We present results for five common benchmark problems: Rastrigin, Sphere,
Schwefel 2.21, Griewank, and Bohachevsky.

Rastrigin is initialized in $[-5.12,5.12]^D$ and defined as:
\[f(\Vec{x}) = \sum_{i=1}^D\left(z_i^2 - 10\cos\left(2\pi z_i\right) +
10\right), z=x-c\]

Sphere is initialized in $[-50,50]^D$ and defined as:
\[f(\Vec{x}) = \sum_{i=1}^D z_i^2, z=x-c\]

Schwefel 2.21 is initialized in $[-500,500]^D$ and defined as:
\[f(\Vec{x}) = \max_i{|z_i|, 1 \leq i \leq D}, z=x-c\]

Griewank is initialized in $[-600,600]^D$ and defined as:
\[f(\Vec{x}) = \frac{1}{4000}\sum_{i=1}^D z_i^2 - \Pi_{i=1}^D
\cos\left(\frac{z_i}{\sqrt{i}} \right) + 1, z=x-c\]

And Bohachevsky is initialized in $[-15,15]^D$ and defined as:
\[f(\Vec{x}) = \sum_{i=1}^D(x_i^2+2x_{i+1}^2-.3\cos(3\pi x_i)-.4\cos(4\pi
x_{i+1})+.7)\]

The $c$ in the function definitions is a shifted center, in order to avoid
origin-seeking bias in the PSO algorithm and its
variants~\cite{monson-2005-origin-seeking-bias}.  We move the center of the
shifted functions (all except Bohachevsky) to be halfway between the center and
the boundary of the initialization region.  For example, if the initialization
region is $[-50,50]^D$, the center is $(25)^D$.

Lastly we look at a function that does not show any kind of improvement with
this method.  The function Rastrigin is defined as 
It has been shown that
with Rastrigin, the more particles there are in the swarm, the lower function
value it finds, up to at least 4000
particles~\cite{mcnabb-2009-large-particle-swarms}.  Smaller swarms get caught
in local optima.  Because our speculative algorithm has a significantly smaller
swarm size, it gets stuck at higher values while the larger swarms performing
regular PSO continue to improve the best value found.  \fig{rastrigin} shows
the results graphically.
\subsection{Comparison with Standard PSO}

The speculative algorithm actually performs two iterations of PSO at each
``iteration,'' so we instead call each ``iteration'' a ``round of
evaluations''.

Where the Complete topology would normally be used, we use a Random topology in
our speculative algorithm, as Complete leads to an explosion in the number of
speculative evaluations.  If speculative evaluation were not being performed,
it is possible that the Complete topology would be used.  However, the Complete
topology also requires a very large amount of interprocessor communication in
distributed PSO, so it is still quite possible that Random would be used even
with standard PSO.  But, to be fair in our comparisons, we compare to the
standard PSO algorithm using both the Random topology with two neighbors and
the Complete topology.  All functions have 20 dimensions, and all results shown
are averages over 20 runs.  We use 240 processors in our experiments (except
where otherwise noted), so our speculative algorithm uses 30 particles and
standard PSO uses 240 particles.  For ease of labeling, we call our speculative
algorithm SEPSO and standard PSO just PSO, with the topology following.

First we look at the function Sphere, the simplest of common benchmark
functions.  Its equation is $f(\Vec{x}) = \sum_{i=1}^D x_i^2$.  The function
has no local optima and is best optimized in terms of function evaluations with
a small swarm using a Complete topology.  Accordingly, we show results for a
Complete swarm in \fig{basic-sphere}.  We can see that when both SEPSO and PSO
are using a Random topology, SEPSO clearly outperforms PSO.  However, the
comparison between SEPSO and PSO Complete is not as encouraging---PSO Complete
performs slightly better.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations.  Our speculative algorithm has a swarm of 30 particles and does
  two iterations per round, while standard PSO algorithms use 240 processors
  and do one iteration per round.}
  \label{fig:basic-sphere}
\end{figure}

@@

\subsection{Pick Best}

We show in \fig{sphere-pickbest} the results of using this modification on the
function Sphere.  For ease of labeling, we call the modified algorithm Pick
Best.  As can be seen, where SEPSO did not manage to perform as well as PSO
Complete, Pick Best handily outperforms it with the same number of processors.

@@

\subsection{Pruning}

There are a large number of ways to prune speculative evaluations.  We
experimented with several, but present results only for one possible pruning.
The pruning we present uses only the two most probable branches in the Random
topology.  As seen in Table~\ref{tab:stats}, these are the two branches where
the $\nbest$ value was not updated: \casepn\ and \casexn.  Those branches are
also convenient in that no messages are needed from neighbors in order to
produce the positions of the speculative particles---in distributed frameworks
using several rounds of communication (see \sec{distributed}), one of the
rounds of communication can be dropped entirely.

Because we are now only doing two speculative evaluations per particle (along
with evaluating the original particle), with 240 processors we can have a swarm
of size 80, instead of the 30 particles in previous experiments.

In \fig{sphere-pruned}, we show the results of this simple pruning on the
function Sphere.  As can be seen, the larger number of particles that pruning
allows leads to better results when picking the best speculative child.  When
using Social Promotion no real improvement is seen over the original PSO
algorithm.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere3}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing pruning with previous results.}
  \label{fig:sphere-pruned}
\end{figure}

@@

\subsection{Many Iters}

The combination of branches we tried uses seven speculative evaluations per
particle, matching the swarm size of the original speculative algorithm.  The
seven evaluations we used corresponded to several iterations of branches
\casepn\ and \casexn.  Two of the evaluations were just one iteration ahead,
four were two iterations ahead, and one was three iterations ahead.  The
evaluations that were one iteration ahead were branches \casepn\ and \casexn;
those that went two iterations ahead were formed by taking either branch
\casepn\ or \casexn and then branch \casepn\ or \casexn; and the evaluation
going three iterations ahead followed branch \casepn\ on all three iterations.

The results when optimizing the function Sphere, presented in
\fig{sphere-manyiters}, are rather astounding.  When 240 processors are
available for use, performing speculative evaluation several iterations ahead
finds values an average of 46 orders of magnitude better than using the
processors to perform the standard PSO algorithm with 240 particles.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere4}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing speculating several iterations ahead with previous
  results.}
  \label{fig:sphere-manyiters}
\end{figure}

@@

>>>>

\section{Experimental Setup}
\label{sec:setup}

The speculative algorithm actually performs two iterations of PSO at each
``iteration,'' so we instead call each ``iteration'' a ``round of
evaluations.''

Where the Complete topology would normally be used, we use a Random topology in
our speculative algorithm, as Complete leads to an explosion in the number of
speculative evaluations.  If speculative evaluation were not being performed,
it is possible that the Complete topology would be used.  However, the Complete
topology also requires a very large amount of interprocessor communication in
distributed PSO, so it is still quite possible that Random would be used even
with standard PSO.  But, to be fair in our comparisons, we compare to the
standard PSO algorithm using both the Random topology with two neighbors and
the Complete topology.

We experimented with five common benchmark problems defined
in~\cite{herrera-2010-test-suite}: Rastrigin, Sphere, Schwefel 2.21, Griewank,
and Bohachevsky.

Rastrigin is initialized in $[-5.12,5.12]^D$ and is defined as:
\[f(\Vec{x}) = \sum_{i=1}^D\left(z_i^2 - 10\cos\left(2\pi z_i\right) +
10\right), z=x-c\]

Sphere is initialized in $[-50,50]^D$ and is defined as:
\[f(\Vec{x}) = \sum_{i=1}^D z_i^2, z=x-c\]

Schwefel 2.21 is initialized in $[-500,500]^D$ and is defined as:
\[f(\Vec{x}) = \max_i{|z_i|, 1 \leq i \leq D}, z=x-c\]

Griewank is initialized in $[-600,600]^D$ and is defined as:
\[f(\Vec{x}) = \frac{1}{4000}\sum_{i=1}^D z_i^2 - \Pi_{i=1}^D
\cos\left(\frac{z_i}{\sqrt{i}} \right) + 1, z=x-c\]

And Bohachevsky is initialized in $[-15,15]^D$ and is defined as:
\[f(\Vec{x}) = \sum_{i=1}^D(x_i^2+2x_{i+1}^2-.3\cos(3\pi x_i)-.4\cos(4\pi
x_{i+1})+.7)\]

The $c$ in the function definitions is a shifted center, in order to avoid
origin-seeking bias in the PSO algorithm and its
variants~\cite{monson-2005-origin-seeking-bias}.  We move the center of the
shifted functions (all except Bohachevsky) to be halfway between the center and
the boundary of the initialization region.  For example, if the initialization
region is $[-50,50]^D$, the center is $(25)^D$.  For all of these functions we
used 20 dimensions.

There are a large number of ways to prune speculative evaluations.  We
experimented with several, but present results for only one possible pruning.
The pruning we present uses only the two most probable branches in the Random
topology.  As seen in Table~\ref{tab:stats}, these are the two branches where
the $\nbest$ value was not updated: \casepn\ and \casexn.  Those branches are
also convenient in that no messages are needed from neighbors in order to
produce the positions of the speculative particles---in distributed frameworks
using several rounds of communication (see \sec{distributed}), one of the
rounds of communication can be dropped entirely.

Because pruning only requires two speculative evaluations per particle (along
with evaluating the original particle), we can use swarms of size $3p$ to
compare to a swarm of size $p$ with standard PSO, instead of $8p$ with other
techniques.

There are also many ways to speculate several iterations ahead, and we again
only show results for one of them.  The combination of branches we tried uses
seven speculative evaluations per particle, matching the swarm size of the
original speculative algorithm.  The seven evaluations we used corresponded to
several iterations of branches \casepn\ and \casexn.  Two of the evaluations
were just one iteration ahead, four were two iterations ahead, and one was
three iterations ahead.  The evaluations that were one iteration ahead were
branches \casepn\ and \casexn; those that went two iterations ahead were formed
by taking either branch \casepn\ or \casexn and then branch \casepn\ or
\casexn; and the evaluation going three iterations ahead followed branch
\casepn\ on all three iterations.

@@

\subsection{Rastrigin}

First we look at Rastrigin.  Rastrigin is a multi-modal function that is best
optimized with a large, Complete swarm.  It has been shown that with Rastrigin,
the more particles there are in the swarm, the lower function value it finds,
up to at least 4000 particles~\cite{mcnabb-2009-large-particle-swarms}.
Smaller swarms get caught in local optima.  Because our speculative algorithms
require significantly smaller swarm sizes, we would expect to not perform very
well on functions such as Rastrigin.  Our experiments show our intuition to be
correct.  In this experiment we used 240 processors, so SEPSO and Pick Best
each had 30 particles, and the standard PSO algorithms had 240 particles.  As
expected, SEPSO and Pick Best converge quickly to worse local optima than
standard PSO does.  \fig{rastrigin} shows the results graphically.

\subsection{Sphere}

Next we go to Sphere, the simplest of common benchmark functions.  The function
has no local optima and is best optimized in terms of function evaluations with
a small swarm using a Complete topology.  We expect our methods to be perfect
for such functions, and our results show this intuition again to be correct.
First we compare the basic SEPSO algorithm to standard PSO.  In
\fig{basic-sphere} we can see that when both SEPSO and PSO are using a Random
topology, SEPSO clearly outperforms PSO.  However, the comparison between SEPSO
and PSO Complete is not as encouraging---PSO Complete performs slightly better.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations.}
  \label{fig:basic-sphere}
\end{figure}

When we introduce Pick Best, we see more impressive results.  As can be seen in
\fig{sphere-pickbest}, where SEPSO did not manage to perform as well as PSO
Complete, Pick Best handily outperforms it with the same number of processors.

Still more impressive with Sphere are the results of pruning and speculating
several iterations ahead.  Pick Best Pruned outperforms Pick Best, and Many
Iters by far outperforms every other algorithm on this function.  Many Iters on
average finds values that are more than 40 orders of magnitude better than
standard PSO using the same number of processors.  This is shown in
\fig{sphere-manyiters}.  Social Promotion Pruned performed comparably with
SEPSO, and we do not clutter the graph with a line for it.

\begin{figure}
  \centering
  \includegraphics[width=.8\columnwidth]{sphere3}
  \caption{Function Sphere with a swarm that uses 240 processors per round of
  evaluations, comparing speculating several iterations ahead with previous
  results.}
  \label{fig:sphere-manyiters}
\end{figure}

\subsection{Schwefel 2.21}

\subsection{Griewank}

@@

@@

@@

@@

\subsection{Bohachevksy}
